---
title: "Compile SDM outputs"
author: "Simon Rolph"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load packages

```{r}
library(dplyr)
library(raster)
```

### Parameters

```{r}
species_name <- "pieris_brassicae"
models <- c("gam","glm","maxent","rf")
AUC_cutoff <- 0.75
```

### LoadAUCs of different models for this species

Get AUCs from models to work out which models to use in the ensemble. Filter by whether they are over the AUC cut-off.

```{r}

#create a dataframe with model type and mean AUC
AUC_df <- data.frame(models,AUC = NA)

#loop through files
for(i in 1:length(models)){
  model <- models[i]
  file_listed <- list.files(paste0("data/derived_data/model_outputs_by_species/",model))
  
  if(paste0("mean_AUC_",species_name,".rds") %in% file_listed){
    AUC_df[i,"AUC"] <- readRDS(paste0("data/derived_data/model_outputs_by_species/",model,"/mean_AUC_",species_name,".rds"))
  }
}

AUC_df

#if there are NA AUCs
if(any(is.na(AUC_df$AUC))){
  AUC_df <- na.omit(AUC_df)
  warning("The AUCs from some models were NA. This may mean that a model failed.")
}

#if any of the models have an AUC over the cut-off
if(any(AUC_df$AUC>AUC_cutoff)){
  #filter so that we only use models over the cut-off
  AUC_df <- AUC_df %>% filter(AUC>AUC_cutoff)
} else{
  #if none are over the threshold then they'll all be used
  warning(paste0("None of the models for ",species_name," met the AUC threshold of ",AUC_cutoff))
}

models_to_load <- AUC_df$models


```

### Load models and produce ensemble

Load models based on the `models_to_load` produced in the previous code block. Use `raster::weighted.mean()` to produce an average weighted by AUC then build and save ensemble model as rasters in .gri/.grd format.

```{r}
# Predicted probability of presence
#load mean model predictions
files_to_load <- paste0("data/derived_data/model_outputs_by_species/",models_to_load,"/mean_prediction_",species_name,".gri")
predictions <- raster::stack(files_to_load)
names(predictions) <- models_to_load

# sd of bootstraps
#load sd of bootstrapped model predictions
files_to_load <- paste0("data/derived_data/model_outputs_by_species/",models_to_load,"/bootstrapped_sd_",species_name,".gri")
bootstrapped_sd <- predictions <- raster::stack(files_to_load)
names(bootstrapped_sd) <- models_to_load

#build ensembles by weighted mean
predictions_ensemble <- raster::weighted.mean(x = predictions, w = AUC_df$AUC)
bootstrapped_sd_ensemble <- raster::weighted.mean(x = bootstrapped_sd, w = AUC_df$AUC)

#save ensembles
raster::writeRaster(predictions_ensemble,
                    filename = paste0("data/derived_data/model_outputs_by_species/ensemble/mean_prediction_",species_name,".grd"),
                    overwrite = TRUE)

raster::writeRaster(bootstrapped_sd_ensemble,
                    filename = paste0("data/derived_data/model_outputs_by_species/ensemble/boostrapped_sd_",species_name,".grd"),
                    overwrite = TRUE)


```


