---
title: "Producing outputs"
author: "UK Centre for Ecology and Hydrology - DECIDE"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#set working directory to the project folder, not the R/scripts folder which it is by default because it's an Rmarkdown doc 
knitr::opts_knit$set(root.dir = '../../')
```


### Load packages

```{r}
library(dplyr)
library(raster)
```

### Parameters and file locations

```{r}
#what group to build the seasonal decide score for
group_to_process <- "butterfly" 
group_to_process <- "day_flying_moth"
group_to_process <- "night_flying_moth" 

#are we accessing the SDMs from the objectstore via datalabs or within the project direction (for when running on JASMIN)
sdm_location <- "datalabs"
#sdm_location  <- "project"


# file locations
## Set up when the app initialises: firstly we need to point to where the raster files are located for each group

if(sdm_location == "datalabs"){
  
  #point to location of all the models
  data_location <- list(
      butterfly = "../../../../data-s3/DECIDE/sdm_outputs/butterfly/combined_model_outputs/PA_thinned_10000nAbs_2021_12_08",
      day_flying_moth = "../../../../data-s3/DECIDE/sdm_outputs/moth/combined_model_outputs/PA_thinned_10000nAbs/mean_score",
      night_flying_moth = "../../../../data-s3/DECIDE/sdm_outputs/nightflying_moth/combined_model_outputs/PA_thinned_10000nAbs_2021_12_08"
  )
  
  #How are the model output files named? presuming that the file names START with the species name
  file_name_template <- list(
      butterfly = list(
          mean = "_PA_thinned_10000nAbs_2021_12_08_mean_score_weightedmeanensemble.grd",
          variation = "_PA_thinned_10000nAbs_2021_12_08_mean_score_weightedvariationensemble.grd",
          auc = "_mean_score_aucOuts.csv"
      ),
      day_flying_moth = list(
          mean = "_PA_thinned_10000nAbs_weightedmeanensemble.grd",
          variation =  "_PA_thinned_10000nAbs_weightedvariationensemble.grd",
          auc = "_mean_score_aucOuts.csv"
      ),
      night_flying_moth = list(
          mean = "_PA_thinned_10000nAbs_2021_12_08_mean_score_weightedmeanensemble.grd",
          variation = "_PA_thinned_10000nAbs_2021_12_08_mean_score_weightedvariationensemble.grd",
          auc = "_mean_score_aucOuts.csv"
      )   
  )
  
} else if(sdm_location == "project") {
  #point to location of all the models
  data_location <- list(
      butterfly = "",
      day_flying_moth = "",
      night_flying_moth = ""
  )
  
  #How are the model output files named?
  file_name_template <- list(
      butterfly = list(
          mean = "",
          variation = "",
          auc = ""
      ),
      day_flying_moth = list(
          mean = "",
          variation =  "",
          auc = ""
      ),
      night_flying_moth = list(
          mean = "",
          variation = "",
          auc = ""
      )   
  )
}

```

### Check files avilability

Taking the species list defined in `species_list.RDS` then building lists of files that we are looking for then comparing this to the files available.

```{r}
#get species list
species_list <- readRDS("species_list.RDS") %>% 
  filter(group == group_to_process) %>% 
  pull(species) %>%
  #tolower() %>% 
  gsub(" ","_",.)


#check ensemble models
#files we want
if(sdm_location == "datalabs"){
  prediction_files <- paste0(species_list,file_name_template[[group_to_process]][["mean"]])
  sd_files <- paste0(species_list,file_name_template[[group_to_process]][["variation"]])
}

if(sdm_location == "project"){
  prediction_files <- paste0("mean_prediction_",species_list,".gri")
  sd_files <- paste0("bootstrapped_sd_",species_list,".gri")
}

file_check_df <- data.frame(species = species_list,
           prediction_file = prediction_files,
           sd_file = sd_files)

#files we have
files_available <- list.files(data_location[[group_to_process]],pattern = ".gri")

#check the target files against files that are available and add a new column to the df reporting on this check
file_check_df$prediction_check <- file_check_df$prediction_file %in% files_available
file_check_df$sd_check <- file_check_df$sd_file %in% files_available

#highlight missing species models but continues none-the-less and makes the outputs with whatever species it has predictions for.
print("Species with predictions:")
species_to_use <- file_check_df %>% filter(prediction_check, prediction_check) %>% dplyr::select(species,prediction_check,sd_check) %>% pull(species)
species_to_use

#species with missing predictions
print("Species with missing predictions:")
file_check_df %>% filter(!prediction_check |!prediction_check) %>% dplyr::select(species,prediction_check,sd_check)

#get the filenames we want to use
prediction_files <- prediction_files[prediction_files %>% grepl(species_to_use,.)]
sd_files <- sd_files[sd_files %>% grepl(species_to_use,.)]

#for testing, slim down the number of files
if(F){
  prediction_files <- prediction_files[1:5]
  sd_files <- sd_files[1:5]
}

```

### Load files and build rasters

Load phenology

```{r}
phenology <- readRDS("data/derived_data/species/phenology.RDS")
```

Load files

```{r}
#prediction raster
prediction_raster <- raster::stack(paste0(data_location[[group_to_process]],prediction_files))
names(prediction_raster) <- species_to_use

#sd raster
sd_raster <- raster::stack(paste0(data_location[[group_to_process]],sd_files))
names(sd_raster) <- species_to_use

```

All-time species richness and uncertainty

```{r}
#build all time richness and model uncertainty layers
species_richness_all_time <- raster::calc(prediction_raster,sum)
model_uncertainty_all_time <- raster::calc(sd_raster,mean)
```

Monthly species richness and uncertainty

```{r}
for (i in 1:12){
  species <- phenology %>% filter(month_num == i,value) %>% pull(scientific_name) %>% tolower() %>% gsub(" ","_",.)
  
  prediction_raster[[names(prediction_raster) %in% species]] %>% raster::calc(sum)
  sd_raster[[names(sd_raster) %in% species]] %>% raster::calc(mean)
  
}


```


Save files

```{r}





```




